A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-16 text/formal-defnition.tex
Add comments on Section 3.3.

We adopt the tradition of \LLF-like systems, where types of constants and kinds of type-level constants are globally declared in a signature $\Sigma$, which
is a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use Boolean in \LMD, $\Sigma$ includes $\B :: *, \textrm{true}:\B, \textrm{false}:\B$.
Note that constants and type-level constants are not associated with stages;
so, we can use them at any stage.
% Type environments
Type environments are sequences of triples of a variable, its type, and its stage.
We define well-formed signature and well-formed type environments later.

@@

In this section, we define the type system of \LMD.
It consists of six judgment forms for typing, kinding, well-formed kinding, term equivalence, type equivalence, and kind equivalence.
We list the judgments forms in Figure~\ref{fig:LMD-six-judgments}.

\begin{figure}
    \begin{center}
        \begin{align*}
            \G & \V M : \tau @ A       & \text{typing}              \\
            \G & \V \tau :: K @ A      & \text{kinding}             \\
            \G & \V K \iskind @ A      & \text{well-formed kinding} \\
            \G & \V M \E N @ A         & \text{term equivalence}    \\
            \G & \V \tau \E \sigma @ A & \text{type equivalence}    \\
            \G & \V K \E J @ A         & \text{kind equivalence}
        \end{align*}
        \caption{Six judgment forms of the type system of \LMD.}
        \label{fig:LMD-six-judgments}
    \end{center}
\end{figure}

\begin{definition}[Typing, Kinding, Well-formed Kinding, Term Equivalence, Type Equivalence, Kind Equivalence]
    The typing relation $ \G \V M : \tau @ A $,
    the kinding relation $\G \V \tau :: K @ A$,
    the well-formed kinding relation $\G \V K \iskind @ A$,
    the term equivalence relation $\G \V M \E N @ A$,
    the type equivalence relation $\G \V \tau \E \sigma @ A$,
    the kind equivalence relation $\G \V K \E J @ A$,
    are the least relation closed under the rules in Figure~\ref{fig:typing-rules}.  \AI{There are more rules!  Some of them are not even in the paper.}
\end{definition}
This definition is little complicated to define six relations at the same time.
We can't define them one by one because they are dependent on each other.
\AI{I think well-formed signatures and type environments cannot be defined independently.}

We introduce a well-formed signature and well-formed type environments.
$\textit{dom}$ is a function from a content to a set of variables defined in the context.
\begin{definition}[Well-formed Signature]
    \begin{itemize}
        \item $\emptyset$ is a well-formed signature.
        \item if $\Sigma$ is a well-formed signature and $\V K :: * @ \varepsilon$ and $X\notin\textit{dom}(\Sigma)$ \\
              then $\Sigma,X::K$ is a well-formed signature.
        \item if $\Sigma$ is a well-formed signature and $\V \tau :: K @ \varepsilon$ and $c\notin\textit{dom}(\Sigma)$ \\
              then $\Sigma,c:\tau$ is a well-formed signature.
    \end{itemize}
\end{definition}
\begin{definition}[Well-formed Type Environments]
    \begin{itemize}
        \item $\emptyset$ are well-formed type environments.
        \item if $\G$ are well-formed type environments and $x\not\in\textit{dom}(\G)$ and $x\not\in\textrm{FV}(\tau)$ \\
              then $\G,x:\tau@A$ are well-formed type environments.
    \end{itemize}
\end{definition}
From here, we assume that all signature and type environments are well-formed.

We start from defining typing of
\LMD.  Typing rules of \LMD are shown in Figure~\ref{fig:typing-rules}.
The rule \TConst{} means any constants in the
signature can appear at any stage.  \
AI{This should be discussed earlier.}
For example, if we have a signature $\Sigma$ which is
$\textrm{bool} :: *, \textrm{true}: \textrm{bool}, \textrm{false}:
    \textrm{bool}$, the derivation tree in
Figure~\ref{fig:tconst-derivation-tree} is admissible.
The rules \TVar,\TAbs, and \TApp{} are almost the same as those in the simply typed
lambda calculus or \LLF.  Additional conditions are that subterms must be
typed at the same stage (\TAbs{} and \TApp); the type
annotation/declaration on a variable has to be a proper type of kind
$*$ (\TAbs) at the stage where it is declared (\TVar{} and \TAbs).
% \AI{Don't we need signature well formedness?  If $c:\tau \in \Sigma$, then $\tau$ should be a well-formed type under ... what?}
% \AI{... or \(\lambda\)LF?  We are working in a dependently type system...}

\begin{figure}
    \begin{center}
        \begin{minipage}{0.50\hsize}
            \infer[\TConst]
            {\G \V \textrm{true}:\textrm{bool}@\alpha\beta}
            {\textrm{true}:\textrm{bool} \in \Sigma \andalso
                \ID{\G\V\textrm{bool}::*@\alpha\beta} \andalso
            }
            \caption{A derivation tree using \TConst}
            \label{fig:tconst-derivation-tree}
        \end{minipage}
    \end{center}
\end{figure}

% \TConv
\AI{Looks like a paper is written for those who know multi-stage calculi fairly well but don't know dependent type systems.  The reality is the opposite.}
As in standard dependent type systems, \TConv{} allows us to replace
the type of a term with an equivalent one.
In a type system which includes dependent types, this kind of rule is essential
because two types which have different shapes may be equivalent.
For example, when we use a vector type with its size ($\textrm{Vector}\ n$),
$\textrm{Vector}\ 5$ is equivalent to $\textrm{Vector}\ (4+1)$ obviously although they are not equivalent in apparently.

% Typing rules for a multi-stage calculus
% \AI{This paragraph is not very informative.  We shouldn't just refer readers to previous work.}
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp{} are rules for a multi-stage calculus.
The rule \TTB{}, which corresponds to brackets, means that if term $M$ is of type $\tau$ at stage $A\alpha$,
$\TB_\alpha M$ is of type $\TW_\alpha \tau$ at stage $A$.
The rule \TTBL{}, which corresponds to escape, is the converse of \TTB.
The rule \TGen{} is a rule for stage abstraction.
The condition of $\alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)$ ensure that
substitutions to $\alpha$ don't affect type environments or stages.
The rule \TIns{} is for applications of stages to stage abstractions.
The rule \TCsp is the rule for CSP, which means that
if term $M$ is of type $\tau$ at stage $A$, $\%_\alpha M$ is of type $\tau$ at stage $A\alpha$.

\begin{figure}
    \begin{center}
        \infrule[\TConst]{c:\tau \in \Sigma \andalso \G\V \tau::*@A}{\G \V c:\tau @A} \\[2mm]
        \infrule[\TVar]{x:\tau @A \in \G \andalso \G\V \tau::*@A}{\G \V x:\tau @A} \\[2mm]
        \infrule[\TAbs]{\G\V \sigma::*@A\andalso\G,x:\sigma@A\V M:\tau @A}{\G\V(\lambda (x:\sigma).M):(\Pi (x:\sigma).\tau)@A} \\[2mm]
        \infrule[\TApp]{\G\V M:(\Pi (x:\sigma).\tau)@A \andalso \G\V N:\sigma@A}{\G\V M\ N : \tau[x\mapsto N]@A} \\[2mm]
        \infrule[\TConv]{\G\V M:\tau @A \andalso \G\V \tau\equiv \sigma :: K@A}{\G\V M:\sigma@A} \\[2mm]
        \infrule[\TTB]{\G\V M:\tau @{A\alpha}}{\G\V\TB_{\alpha}M:\TW_{\alpha}\tau @A} \andalso
        \infrule[\TTBL]{\G\V M:\TW_{\alpha}\tau @A}{\G\V\TBL_{\alpha}M:\tau @{A\alpha}} \\[2mm]
        \infrule[\TGen]{\G\V M:\tau @A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V\Lambda\alpha.M:\forall\alpha.\tau @A} \\[2mm]
        \infrule[\TIns]{\G\V M:\forall\alpha.\tau @A}{\G\V M\ B:\tau[\alpha \mapsto B]@A} \andalso
        \infrule[\TCsp]{\G\V M:\tau @A}{\G\V \%_\alpha M:\tau @{A\alpha}}
        \caption{Typing Rules.}
        \label{fig:typing-rules}
    \end{center}
\end{figure}

\subsubsection{Type and Term Equivalence}

As usual in dependent type systems, type equivalence, used in \TConv,
is important.  The type equivalence judgment of the form
$\G \V \tau \E \sigma : K @ A$ means that types $\tau$ and $\sigma$
are equivalent as types of kind $K$ at stage $A$ under $\G$.
Figure~\ref{fig:type-equivalence-rules} shows the rules for type
equivalence.  Type equivalence is basically the least congruence
closed under term equivalence.  The rules for compatibility (closure
under type formation) are derived from corresponding typing rules in a
straightforward manner.  The rules are a little simpler than some
dependent type systems \AI{such as?}  because there is no abstraction
at the type level.

@@

\begin{figure}
    \begin{center}
        \infrule[{\QTAbs }]{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau @A \V \rho \E \pi :: *@A}{\G\V\Pi x:\tau.\rho \E \Pi x:\sigma.\pi :: *@A} \\[2mm]
        \infrule[\QTApp]{\G\V \tau \E \sigma :: (\Pi x:\rho.K)@A \andalso \G\V M \E N : \rho @A}{\G\V \tau\ M \E \sigma\ N :: K[x \mapsto M]@A} \\[2mm]
        \infrule[\textsc{QT-$\TW$}]{\G\V \tau \E \sigma :: *@{A\alpha}}{\G\V \TW_{\alpha} \tau \E \TW_{\alpha} \sigma :: *@A}\andalso
        \infrule[\QTCsp]{\G\V \tau \E \sigma :: K@A}{\G\V \tau \E \sigma :: K@{A\alpha}} \\[2mm]
        \infrule[\QTGen]{\G\V \tau \E \sigma :: *@A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V \forall\alpha.\tau \E  \forall\alpha.\sigma :: *@A} \\[2mm]
        \infrule[\QTRefl]{\G\V \tau::K@A}{\G\V \tau\E\tau :: K@A} \andalso
        \infrule[\QTSym]{\G\V \tau \E \sigma :: K@A}{\G\V \sigma \E \tau :: K@A} \\[2mm]
        \infrule[\QTTrans]{\G\V \tau \E \sigma :: K@A \andalso \G\V \sigma \E \rho  :: K@A}{\G\V \tau \E \rho  :: K@A}
        \caption{Type Equivalence Rules.}
        \label{fig:type-equivalence-rules}
    \end{center}
\end{figure}

The term equivalence judgment of the form $\G \V M \E N : \rho @ A$,
which means that terms $M$ and $N$ are equivalent as terms of type
$\rho$ at stage $A$ under $\G$, is defined by the rules in
Figure~\ref{fig:term-equivalence-rules}.  Most rules are
straightforward.  The rules \QAbs, \QApp, \QTB, \QTBL, \QGen, \QIns,
\QCsp, \QRefl, \QSym, and \QTrans{} make the relation congruence; the
rules \QBeta, \QTBLTB, and \QLambda{} correspond to
$\beta$-reduction, $\blacklozenge$-reduction, and $\Lambda$-reduction, respectively.

\begin{figure}
    \begin{center}
        \infrule[\QAbs]{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau @A \V M \E N : \rho @A}{\G\V\lambda x:\tau.M \E \lambda x:\sigma.N : (\Pi x:\tau.\rho)@A} \\[2mm]
        \infrule[\QAbs]{\G\V M \E L : (\Pi x:\sigma.\tau)@A \andalso \G\V N \E O : \sigma@A}{\G\V M\ N \E L\ O : \tau[x \mapsto N]@A} \\[2mm]
        \infrule[\QTB]{\G\V M \E N : \tau @{A\alpha}}{\G\V \TB_\alpha M \E \TB_\alpha N : \TW_\alpha \tau @A} \andalso
        \infrule[\QTBL]{\G\V M \E N : \TW_\alpha \tau @A}{\G\V \TBL_\alpha M \E \TBL_\alpha N : \tau @{A\alpha}} \\[2mm]
        \infrule[\QGen]{\G\V M\E N : \tau @A \andalso \alpha \notin \FTV(\G)\cup\FTV(A)}{\G\V \Lambda\alpha.M \E \Lambda\alpha.N : \forall\alpha.\tau @A} \\[2mm]
        \infrule[\QIns]{\G\V M \E N:\forall\alpha.\tau @A}{\G\V M\ \varepsilon \E N\ \varepsilon : \tau[\alpha \mapsto \varepsilon]@A} \andalso
        \infrule[\QCsp]{\G\V M \E N : \tau @A}{\G\V\%_\alpha M \E \%_\alpha N : \tau @{A\alpha}} \\[2mm]
        \infrule[\QRefl]{\G\V M:\tau @A}{\G\V M\E M : \tau @A} \andalso
        \infrule[\QSym]{\G\V M\E N : \tau @A}{\G\V N\E M : \tau @A} \\[2mm]
        \infrule[\QTrans]{\G\V M\E N : \tau @A \andalso \G\V N\E L : \tau @A}{\G\V M\E L : \tau @A} \\[2mm]
        \infrule[\QBeta]{\G,x:\sigma@A\V M:\tau @A \andalso \G\V N:\sigma@A}{\G\V(\lambda x:\sigma.M)\ N\E M[x\mapsto N] : \tau[x \mapsto N]@A} \\[2mm]
        % \infrule{\G\V M:(\Pi x:\sigma.\tau)@A \andalso x\notin \text{FV}(M)}{\G\V(\lambda x:\sigma.M\ x)\E M: (\Pi x:\sigma.\tau)@A}{\QEta} \\[2mm]
        \infrule[\QLambda]{\G\V (\Lambda\alpha.M) : \forall\alpha.\tau @A}{\G\V (\Lambda\alpha.M)\ \varepsilon \E M[\alpha \mapsto \varepsilon] : \tau[\alpha \mapsto \varepsilon]@A} \\[2mm]
        \infrule[\QTBLTB]{\G\V M \E N : \tau @A}{\G\V \TBL_\alpha(\TB_\alpha M) \E N : \tau @A} \hfil
        \infrule[\QPercent]{\G\V M:\tau @{A\alpha} \andalso \G\V M:\tau @A}{\G\V\%_\alpha M \E M : \tau @{A\alpha}}
        \caption{Term Equivalence Rules.}
        \label{fig:term-equivalence-rules}
    \end{center}
\end{figure}

@@

Interestingly, Hanada and Igarashi~\cite{Hanada2014} rejected the idea of
reduction to remove $\%_\alpha$ when they developed \LTP{}, as such
the reduction does not match the operational behavior of the CSP operator
in implementation.  However, as an equational system for multi-stage
programs, the rule \QPercent makes sense and as we argue next
it is practically significant.

\renewcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
As we discussed in Section 2, $\text{vadd}$ function, 
which takes the length of vectors and returns a quoted add function for vector of the given size,
has the type of $\Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
When we use this function, we instansiate $\text{vadd}$ with some integer such as 5
and get code of a optimzed $\text{vadd}$ function such as $\text{vadd5}$.
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 5)}
Although $\text{vadd5}$ is of type $\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$,
a code value of this type is difficult to combine with another code fragment.
For example, if we want to composite $\text{vadd5}$ with reverse fucntion, 
which is of type $\text{rev5}:\F\alpha.\TW_\alpha(\text{Vector}\ 5\to\text{Vector}\ 5)$,
we write a program like following.
\begin{align*}
\Lambda\gamma.\TB_\gamma.\lambda v_1:(\text{Vector}\ 5).\lambda v_2:(\text{Vector}\ 5).\TBL_\gamma 
(\text{rev5}\ \gamma)\ (\TBL_\gamma (\text{vadd5}\ \gamma)\ \TBL_\gamma v_1\ \TBL_\gamma v_2)
\end{align*}
However, this program cannot be typed without \QPercent{}
because the type of $v_1$: $\text{Vector}\ 5$ isn't equivalent with the type of arguments of $\text{vadd5}$: $\text{Vector}\ \%_\gamma 5$.

@@

\subsubsection{Staged Semantics}

@@

In order to define Staged Reduction, we define redex and evaluation context.

\begin{definition}[Redex]
    the sets of $\varepsilon$-redexes (ranged over by $R^\varepsilon$) and $\alpha$-redexes (ranged over by $R^\alpha$) are defined by the following grammar.
    \begin{align*}
         & R^\varepsilon ::= (\lambda x:\tau.M)\ v^\varepsilon \mid (\Lambda\alpha.v^\varepsilon)\ \varepsilon \\
         & R^\alpha      ::=\ \TBL_\alpha \TB_\alpha M                                                         \\
    \end{align*}
\end{definition}

>>>>

We adopt the tradition of \LLF-like systems, where types of constants and kinds of type-level constants are globally declared in a signature $\Sigma$, which
is a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use Boolean in \LMD, $\Sigma$ includes $\B :: *, \textrm{true}:\B, \textrm{false}:\B$.
Type environments are sequences of triples of a variable, its type, and its stage.
We write \(\textit{dom}(\Sigma)\) and \(\textit{dom}(\Gamma)\)
for the set of (type-level) constants declared in \(\Sigma\) and \(\Gamma\), respectively.   As in other multi-stage calculi~\cite{Taha03,Tsukada,Hanada2014},
a variable declaration is associated with a stage
so that a variable can be referenced only at the declared stage.
Constants and type-level constants are not associated with stages;
so, we can use them at any stage.
We define well-formed signature and well-formed type environments later.

@@

In this section, we define the type system of \LMD.
It consists of eight judgment forms for signature well-formedness, type environment well-formedness, kind well-formedness, kinding, typing, term equivalence, type equivalence, and kind equivalence.
We list the judgments forms in Figure~\ref{fig:LMD-six-judgments}.
They are all defined in a mutual recursive manner.  We will discuss
each judgment below.

\begin{figure}
    \begin{center}
      \begin{align*}
            \vdash & \Sigma            & \text{signature well-formedness} \\
            \V & \G                & \text{type environment well-formedness} \\
            \G & \V K \iskind @ A      & \text{kind well-formedness} \\
            \G & \V \tau :: K @ A      & \text{kinding}             \\
            \G & \V M : \tau @ A       & \text{typing}              \\
            \G & \V K \E J @ A         & \text{kind equivalence}    \\
            \G & \V \tau \E \sigma @ A & \text{type equivalence}    \\
            \G & \V M \E N @ A         & \text{term equivalence}
        \end{align*}
        \caption{Eight judgment forms of the type system of \LMD.}
        \label{fig:LMD-six-judgments}
    \end{center}
\end{figure}

\subsubsection{Signature and Type Environment Well-formedness.}
The rules for Well-formed signatures and type environments are
shown below:
%
{\small
\begin{center}
  \infrule{
  }{
    \vdash \emptyset
  }
  \hfil
  \infrule{
    \vdash \Sigma \andalso
    \V K \iskind @ \varepsilon \\
    X\notin\textit{dom}(\Sigma)
  }{
    \vdash \Sigma, X::K
  }
  \hfil
  \infrule{
    \vdash \Sigma \andalso
    \V \tau :: * @ \varepsilon \\
    c\notin\textit{dom}(\Sigma)
  }{
   \vdash \Sigma, c:\tau
 }
 \\[2mm]
  \infrule{
  }{
    \V \emptyset
  }
  \hfil
  \infrule{
    \V \Gamma \andalso
    \Gamma \V \tau :: * @ A \andalso
    x\notin\textit{dom}(\Sigma)
  }{
   \V \Gamma, x:\tau@A
 }
\end{center}
}

To add declarations to a signature, the kind/type of a (type-level)
constant has to be well-formed at stage \(\varepsilon\) so that it is
used at any stage. \AI{We should discuss implicit CSP at the type
  level somewhere.}

In what follows, well-formedness is not explicitly mentioned but
we assume that all signatures and type environments are well-formed.

\subsubsection{Kind Wellformedness and Kinding.}

The rules for kind wellformedness and kiding is straightforward
adaptation from \LLF.  So, we omit them.

\subsubsection{Typing.}

\AI{The second premises of \TConst and \TVar are not needed?}
\begin{figure}
    \begin{center}
        \infrule[\TConst]{c:\tau \in \Sigma \andalso \G\V \tau::*@A}{\G \V c:\tau @A} \hfil
        \infrule[\TVar]{x:\tau @A \in \G \andalso \G\V \tau::*@A}{\G \V x:\tau @A} \\[2mm]
        \infrule[\TAbs]{\G\V \sigma::*@A\andalso\G,x:\sigma@A\V M:\tau @A}{\G\V(\lambda (x:\sigma).M):(\Pi (x:\sigma).\tau)@A} \\[2mm]
        \infrule[\TApp]{\G\V M:(\Pi (x:\sigma).\tau)@A \andalso \G\V N:\sigma@A}{\G\V M\ N : \tau[x\mapsto N]@A} \\[2mm]
        \infrule[\TConv]{\G\V M:\tau @A \andalso \G\V \tau\equiv \sigma :: K@A}{\G\V M:\sigma@A} \\[2mm]
        \infrule[\TTB]{\G\V M:\tau @{A\alpha}}{\G\V\TB_{\alpha}M:\TW_{\alpha}\tau @A} \andalso
        \infrule[\TTBL]{\G\V M:\TW_{\alpha}\tau @A}{\G\V\TBL_{\alpha}M:\tau @{A\alpha}} \\[2mm]
        \infrule[\TGen]{\G\V M:\tau @A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V\Lambda\alpha.M:\forall\alpha.\tau @A} \\[2mm]
        \infrule[\TIns]{\G\V M:\forall\alpha.\tau @A}{\G\V M\ B:\tau[\alpha \mapsto B]@A} \andalso
        \infrule[\TCsp]{\G\V M:\tau @A}{\G\V \%_\alpha M:\tau @{A\alpha}}
        \caption{Typing Rules.}
        \label{fig:typing-rules}
    \end{center}
\end{figure}

Typing rules of \LMD are shown in Figure~\ref{fig:typing-rules}.
The rule \TConst{} means that a constant can appear at any stage.
% For example, if we have a signature $\Sigma$ which is
% $\textrm{bool} :: *, \textrm{true}: \textrm{bool}, \textrm{false}:
%     \textrm{bool}$, the derivation tree in
% Figure~\ref{fig:tconst-derivation-tree} is admissible.
The rules \TVar,\TAbs, and \TApp{} are almost the same as those in the simply typed
lambda calculus or \LLF.  Additional conditions are that subterms must be
typed at the same stage (\TAbs{} and \TApp); the type
annotation/declaration on a variable has to be a proper type of kind
$*$ (\TAbs) at the stage where it is declared (\TVar{} and \TAbs).
% \AI{Don't we need signature well formedness?  If $c:\tau \in \Sigma$, then $\tau$ should be a well-formed type under ... what?}
% \AI{... or \(\lambda\)LF?  We are working in a dependently type system...}

% \begin{figure}
%     \begin{center}
%         \begin{minipage}{0.50\hsize}
%             \infer[\TConst]
%             {\G \V \textrm{true}:\textrm{bool}@\alpha\beta}
%             {\textrm{true}:\textrm{bool} \in \Sigma \andalso
%                 \ID{\G\V\textrm{bool}::*@\alpha\beta} \andalso
%             }
%             \caption{A derivation tree using \TConst}
%             \label{fig:tconst-derivation-tree}
%         \end{minipage}
%     \end{center}
% \end{figure}

% \TConv
\AI{Looks like a paper is written for those who know multi-stage calculi fairly well but don't know dependent type systems.  The reality is the opposite.}
As in standard dependent type systems, \TConv{} allows us to replace
the type of a term with an equivalent one.
In a type system with dependent types, this kind of rule is essential
because \AI{?} two types which have different shapes may be equivalent.
For example, when we use a vector type with its size ($\textrm{Vector}\ n$),
$\textrm{Vector}\ 5$ is equivalent to $\textrm{Vector}\ (4+1)$ obviously although they are not syntactically equivalent.

% Typing rules for a multi-stage calculus
% \AI{This paragraph is not very informative.  We shouldn't just refer readers to previous work.}
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp{} are rules for a multi-stage calculus.
The rule \TTB{}, which corresponds to brackets \AI{Sounds like the rule corresponds to brackets.  What corresponds to brackets is $\TB$}, means that if term $M$ is of type $\tau$ at stage $A\alpha$,
$\TB_\alpha M$ is of type $\TW_\alpha \tau$ at stage $A$.
The rule \TTBL{}, which corresponds to escape, is the converse of \TTB.
The rule \TGen{} for stage abstraction is straightforward.
The condition $\alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)$ ensures that
the scope of $\alpha$ is in $M$.
The rule \TIns{} is for applications of stages to stage abstractions.
\AI{Any comments?}
The rule \TCsp is the rule for CSP, which means that
if term $M$ is of type $\tau$ at stage $A$, $\%_\alpha M$ is of type $\tau$ at stage $A\alpha$.  \AI{This does not explain what this rule means.
  It just gives how the rule is read.}

\subsubsection{Kind, Type and Term Equivalence.}

Since the syntax of kinds, types, and terms is mutually recursive,
corresponding notions of equivalence are also mutually recursive.
They are congruences closed under a few axioms for term equivalence.
So, the rules for kind and type equivalences are not very interesting
except implict CSP is allowed.
We show a few representative rules below.

{\small
\begin{center}
\infrule[\textsc{QK-Csp}]{%
  \G\V K \E J @ A
}{
  \G\V K \E J @ A\alpha
}
\hfil
\infrule[\QTCsp]{
  \G\V \tau \E \sigma :: K@A
}{
  \G\V \tau \E \sigma :: K@{A\alpha}
}
\\[2mm]
\infrule[\QTApp]{%
  \G\V \tau \E \sigma :: (\Pi x:\rho.K)@A \andalso
  \G\V M \E N : \rho @A
}{
  \G\V \tau\ M \E \sigma\ N :: K[x \mapsto M]@A
}
\end{center}
}

% The type equivalence judgment of the form
% $\G \V \tau \E \sigma : K @ A$ means that types $\tau$ and $\sigma$
% are equivalent as types of kind $K$ at stage $A$ under $\G$.
% Figure~\ref{fig:type-equivalence-rules} shows the rules for type
% equivalence.  Type equivalence is basically the least congruence
% closed under term equivalence.  The rules for compatibility (closure
% under type formation) are derived from corresponding typing rules in a
% straightforward manner.  The rules are a little simpler than some
% dependent type systems \AI{such as?}  because there is no abstraction
% at the type level.

@@

% \begin{figure}
%     \begin{center}
%         \infrule[{\QTAbs }]{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau @A \V \rho \E \pi :: *@A}{\G\V\Pi x:\tau.\rho \E \Pi x:\sigma.\pi :: *@A} \\[2mm]
%         \infrule[\QTApp]{\G\V \tau \E \sigma :: (\Pi x:\rho.K)@A \andalso \G\V M \E N : \rho @A}{\G\V \tau\ M \E \sigma\ N :: K[x \mapsto M]@A} \\[2mm]
%         \infrule[\textsc{QT-$\TW$}]{\G\V \tau \E \sigma :: *@{A\alpha}}{\G\V \TW_{\alpha} \tau \E \TW_{\alpha} \sigma :: *@A}\andalso
%         \infrule[\QTCsp]{\G\V \tau \E \sigma :: K@A}{\G\V \tau \E \sigma :: K@{A\alpha}} \\[2mm]
%         \infrule[\QTGen]{\G\V \tau \E \sigma :: *@A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V \forall\alpha.\tau \E  \forall\alpha.\sigma :: *@A} \\[2mm]
%         \infrule[\QTRefl]{\G\V \tau::K@A}{\G\V \tau\E\tau :: K@A} \andalso
%         \infrule[\QTSym]{\G\V \tau \E \sigma :: K@A}{\G\V \sigma \E \tau :: K@A} \\[2mm]
%         \infrule[\QTTrans]{\G\V \tau \E \sigma :: K@A \andalso \G\V \sigma \E \rho  :: K@A}{\G\V \tau \E \rho  :: K@A}
%         \caption{Type Equivalence Rules.}
%         \label{fig:type-equivalence-rules}
%     \end{center}
% \end{figure}

% The term equivalence judgment of the form $\G \V M \E N : \rho @ A$,
% which means that terms $M$ and $N$ are equivalent as terms of type
% $\rho$ at stage $A$ under $\G$, is defined by the rules in
We show the rules for term equivalence are shown in 
Figure~\ref{fig:term-equivalence-rules}, omitting
straightforward rules for reflexivity, symmetry, transitivity,
and compatibility.
% Most rules are
% straightforward.  The rules \QAbs, \QApp, \QTB, \QTBL, \QGen, \QIns,
% \QCsp, \QRefl, \QSym, and \QTrans{} make the relation congruence;
The rules \QBeta, \QTBLTB, and \QLambda{} correspond to
$\beta$-reduction, $\blacklozenge$-reduction, and $\Lambda$-reduction, respectively.

\begin{figure}
    \begin{center}
        % \infrule[\QAbs]{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau @A \V M \E N : \rho @A}{\G\V\lambda x:\tau.M \E \lambda x:\sigma.N : (\Pi x:\tau.\rho)@A} \\[2mm]
        % \infrule[\QApp]{\G\V M \E L : (\Pi x:\sigma.\tau)@A \andalso \G\V N \E O : \sigma@A}{\G\V M\ N \E L\ O : \tau[x \mapsto N]@A} \\[2mm]
        % \infrule[\QTB]{\G\V M \E N : \tau @{A\alpha}}{\G\V \TB_\alpha M \E \TB_\alpha N : \TW_\alpha \tau @A} \andalso
        % \infrule[\QTBL]{\G\V M \E N : \TW_\alpha \tau @A}{\G\V \TBL_\alpha M \E \TBL_\alpha N : \tau @{A\alpha}} \\[2mm]
        % \infrule[\QGen]{\G\V M\E N : \tau @A \andalso \alpha \notin \FTV(\G)\cup\FTV(A)}{\G\V \Lambda\alpha.M \E \Lambda\alpha.N : \forall\alpha.\tau @A} \\[2mm]
        % \infrule[\QIns]{\G\V M \E N:\forall\alpha.\tau @A}{\G\V M\ \varepsilon \E N\ \varepsilon : \tau[\alpha \mapsto \varepsilon]@A} \andalso
        % \infrule[\QCsp]{\G\V M \E N : \tau @A}{\G\V\%_\alpha M \E \%_\alpha N : \tau @{A\alpha}} \\[2mm]
        % \infrule[\QRefl]{\G\V M:\tau @A}{\G\V M\E M : \tau @A} \andalso
        % \infrule[\QSym]{\G\V M\E N : \tau @A}{\G\V N\E M : \tau @A} \\[2mm]
        % \infrule[\QTrans]{\G\V M\E N : \tau @A \andalso \G\V N\E L : \tau @A}{\G\V M\E L : \tau @A} \\[2mm]
        \infrule[\QBeta]{\G,x:\sigma@A\V M:\tau @A \andalso \G\V N:\sigma@A}{\G\V(\lambda x:\sigma.M)\ N\E M[x\mapsto N] : \tau[x \mapsto N]@A} \\[2mm]
        % \infrule{\G\V M:(\Pi x:\sigma.\tau)@A \andalso x\notin \text{FV}(M)}{\G\V(\lambda x:\sigma.M\ x)\E M: (\Pi x:\sigma.\tau)@A}{\QEta} \\[2mm]
        \infrule[\QLambda]{\G\V (\Lambda\alpha.M) : \forall\alpha.\tau @A}{\G\V (\Lambda\alpha.M)\ \varepsilon \E M[\alpha \mapsto \varepsilon] : \tau[\alpha \mapsto \varepsilon]@A} \\[2mm]
        \infrule[\QTBLTB]{\G\V M \E N : \tau @A}{\G\V \TBL_\alpha(\TB_\alpha M) \E N : \tau @A} \hfil
        \infrule[\QPercent]{\G\V M:\tau @{A\alpha} \andalso \G\V M:\tau @A}{\G\V\%_\alpha M \E M : \tau @{A\alpha}}
        \caption{Term Equivalence Rules.}
        \label{fig:term-equivalence-rules}
    \end{center}
\end{figure}

@@

Interestingly, Hanada and Igarashi~\cite{Hanada2014} rejected the idea of
reduction that removes $\%_\alpha$ when they developed \LTP{}, as such
reduction does not match the operational behavior of the CSP operator
in implementation.  However, as an equational system for multi-stage
programs, the rule \QPercent makes sense and, as we argue next,
it is practically significant.

\AI{This paragraph should be polished later.}
\renewcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
As we discussed in Section 2, $\text{vadd}$ function, 
which takes the length of vectors and returns a quoted add function for vector of the given size,
has the type of $\Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
When we use this function, we instansiate $\text{vadd}$ with some integer such as 5
and get code of a optimzed $\text{vadd}$ function such as $\text{vadd5}$.
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 5)}
Although $\text{vadd5}$ is of type $\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$,
a code value of this type is difficult to combine with another code fragment.
For example, if we want to composite $\text{vadd5}$ with reverse fucntion, 
which is of type $\text{rev5}:\F\alpha.\TW_\alpha(\text{Vector}\ 5\to\text{Vector}\ 5)$,
we write a program like following.
\begin{align*}
\Lambda\gamma.\TB_\gamma.\lambda v_1:(\text{Vector}\ 5).\lambda v_2:(\text{Vector}\ 5).\TBL_\gamma 
(\text{rev5}\ \gamma)\ (\TBL_\gamma (\text{vadd5}\ \gamma)\ \TBL_\gamma v_1\ \TBL_\gamma v_2)
\end{align*}
However, this program cannot be typed without \QPercent{}
because the type of $v_1$: $\text{Vector}\ 5$ isn't equivalent with the type of arguments of $\text{vadd5}$: $\text{Vector}\ \%_\gamma 5$.

@@

\subsubsection{Staged Semantics.}

@@

In order to define staged reduction, we define redex and evaluation contexts.

\begin{definition}[Redex]
    The sets of $\varepsilon$-redexes (ranged over by $R^\varepsilon$) and $\alpha$-redexes (ranged over by $R^\alpha$) are defined by the following grammar.
    \begin{align*}
         & R^\varepsilon ::= (\lambda x:\tau.M)\ v^\varepsilon \mid (\Lambda\alpha.v^\varepsilon)\ \varepsilon \\
         & R^\alpha      ::=\ \TBL_\alpha \TB_\alpha M                                                         \\
    \end{align*}
\end{definition}
